A Survey of Quantitative Descriptions of Molecular Structure

### 2012-07-04 paper.tex
A few spelling fixes

Geometric descriptors require a 3D conformation as input and therefore
the computations are not as fast as topological descriptors which do
not require the 3D coordinate generation step. Nearly all geometric
descriptors work on a single conformation and for cases where multiple
conformation must be considered, an averaging procedure is usually
employed. Examples of these descriptors include the gravitational
index \cite{Katritzky:1996ly} and moment of inertia descriptors.
Geometric descriptors encompass many aspects of a molecular
structure. For example, a number of them characterize the molecular
surface. The simplest such example is the surface area descriptor (and
the corresponding volume descriptor). A simple extension of such a
descriptor is to characterize distributions of a physicochemical
property over the molecular surface such as the Charged Partial
Surface Area descriptors \cite{Stanton:1990aa} that characterize the
partial charge distribution over the molecular surfce. Note that
surface descriptor calculations based on analytical
\cite{Connolly:1983aa} or tesselation algorithms
\cite{Eisenhaber:1995qf} can be slow and that parametrized methods
such as the Topological Polar Surface Area (TPSA) \cite{Ertl:2000aa}
and VSA descriptors \cite{Labute:2008aa} that are based on
precalculated surface area values derived from a list of functional
groups.

@@

Another important class of geometrical descriptors are those that
describe molecular shape. This property is a key factor in ligand-receptor interactions and is
an important approach to virtual screening. A variety of descriptors
have been devised to characterize molecular shape. The use of
overlapping atom centered Gaussians to represent molecular shape
\cite{Grant:1995aa} represents a significant computational advantage
over the traditional analytical approach of intersecting hard spheres,
supporting extremely rapid shape comparisons \cite{Grant:1999vn}. An
alternative approach described by Ritchie et al \cite{Ritchie:1999kx}
employed a spherical harmonic based representation of surfaces, that
in contrast to the Gaussian approach allows one to generate
characterizations of the surface and hence the shape at varying
resolutions. Another approach to multi-resolution shape
representations is the use of $\alpha$ shapes
\cite{Edelsbrunner:1994aa}, which represent a parametrization of the
convex hull derived from the molecular coordinates. This approach has
been employed to characterize shape and volume \cite{Wilson:2009ys}
though it is more computationally intensive than the methods mentioned
previously. In addition to functional representations of molecular
shape, a number of simpler formulations have been defined. For
example, the ``shape signature'' approach described by Zauhar et al
\cite{Zauhar:2003fk} is a novel application of a ray tracing method
that generates a description of the molecular shape. Ballester et al
\cite{Ballester:2007aa} described the Ultrafast Shape Recognition
(USR) approach that characterized molecular shape in terms of the
distribution of distances between the atoms and a set of pre-defined
points. This procedure characterizes the shape in the form of a 12-D
real valued fingerprint. As a result, the calculation procedure is
rapid and once calculated shape similarities can be evaluated rapidly
via a simple distance calculation. It should also be noted that shape
indices based on the molecular topology have also been described
\cite{Kier:1985aa}, which are very rapid to calculate (somewhat
analogous to the VSA descriptors) but do not directly represent the
molecular shape and instead infer it from topological connectivity.

@@

\textbf{Descriptors for inorganic compounds} Inorganic materials are
sufficiently different from small molecules that traditional
descriptors are not very useful. Recently, a number of descriptors for
inorganic materials are very similar to constitutional descriptors -
essentially counts of various crystal parameters such as accessible
surface areas and volumes
\cite{Willems:2012zr,Haranczyk:2010ys,Mackay:1984ve}. Other
descriptors characterize the periodic voids of porous materials via
tesselation methods \cite{Anurova:2009ly,Carr:2009kx}.

@@

The requirements listed here ensures that subsequent analyses
involving descriptors are robust. But in addition to numerical
robustness, it is important for many scenarios that the descriptor
have some form of physical interpretability.  For example, a model
that employs descriptors to predict cytoxicity will be most useful if
it is numerically accurate as well as providing insight into which
structural features confer toxicity. In such a scenario, abstract
descriptors do not allow us to provide a physical explanation of the
models predictions. Another scenario is the ability to map descriptor
values back to the structure for visualization purposes such as the
``Glowing Molecule'' depictions of Segall et al
\cite{Segall:2009fk}. Such visualizations are practical only when
descriptor values can be related to structural features. Examples
include surface based descriptors such as CPSA's as well as explicit
structural descriptors such as paths, chains etc.

@@

Given the multiple implementations of many descriptors, it is natural
to compare their performance. As noted previously, it is desirable,
but unlikely, that implementations of the same descriptor using
different cheminformatcs toolkits will be identical. Gupta et al
employed SMARTS based descriptors from the CDK and MOE to develop
decision tree models to predict human liver microsomal metabolic
stability \cite{Gupta:2010uq}. Their results indicated very similar
performance between the two implementations. In general, one would
expect a high degree of correlation between different implementations
of well-defined descriptors. Here ``well defined'' indicates that the
underlying parametrization or chemistry model is completely specified,
so that different implementations use them identically. For example,
correct implementations of the TPSA descriptor are expected to
correlate very well, and this is shown in Figure \ref{fig:tpsa} for
the ACD Labs and CDK implementations.

@@

\ctable[caption={A comparison of Topological Polar Surface Area values
generated using the CDK and ACD Labs software, for 57,857 molecules
taken from Pubchem AID 1996.},cap={},label={fig:tpsa},botcap,figure]
{c}
{}
{
  \includegraphics[width=\linewidth]{tpsa-cdk-acd}
}

>>>>

Geometric descriptors require a 3D conformation as input and therefore
the computations are not as fast as topological descriptors which do
not require the 3D coordinate generation step. Nearly all geometric
descriptors work on a single conformation and for cases where multiple
conformation must be considered, an averaging procedure is usually
employed. Examples of these descriptors include the gravitational
index \cite{Katritzky:1996ly} and moment of inertia descriptors.
Geometric descriptors encompass many aspects of a molecular
structure. For example, a number of them characterize the molecular
surface. The simplest such example is the surface area descriptor (and
the corresponding volume descriptor). A simple extension of such a
descriptor is to characterize distributions of a physicochemical
property over the molecular surface such as the Charged Partial
Surface Area descriptors \cite{Stanton:1990aa} that characterize the
partial charge distribution over the molecular surface. Note that
surface descriptor calculations based on analytical
\cite{Connolly:1983aa} or tessellation algorithms
\cite{Eisenhaber:1995qf} can be slow and that parameterized methods
such as the Topological Polar Surface Area (TPSA) \cite{Ertl:2000aa}
and VSA descriptors \cite{Labute:2008aa} that are based on
precalculated surface area values derived from a list of functional
groups.

@@

Another important class of geometrical descriptors are those that
describe molecular shape. This property is a key factor in ligand-receptor interactions and is
an important approach to virtual screening. A variety of descriptors
have been devised to characterize molecular shape. The use of
overlapping atom centered Gaussians to represent molecular shape
\cite{Grant:1995aa} represents a significant computational advantage
over the traditional analytical approach of intersecting hard spheres,
supporting extremely rapid shape comparisons \cite{Grant:1999vn}. An
alternative approach described by Ritchie et al \cite{Ritchie:1999kx}
employed a spherical harmonic based representation of surfaces, that
in contrast to the Gaussian approach allows one to generate
characterizations of the surface and hence the shape at varying
resolutions. Another approach to multi-resolution shape
representations is the use of $\alpha$ shapes
\cite{Edelsbrunner:1994aa}, which represent a parameterization of the
convex hull derived from the molecular coordinates. This approach has
been employed to characterize shape and volume \cite{Wilson:2009ys}
though it is more computationally intensive than the methods mentioned
previously. In addition to functional representations of molecular
shape, a number of simpler formulations have been defined. For
example, the ``shape signature'' approach described by Zauhar et al
\cite{Zauhar:2003fk} is a novel application of a ray tracing method
that generates a description of the molecular shape. Ballester et al
\cite{Ballester:2007aa} described the Ultrafast Shape Recognition
(USR) approach that characterized molecular shape in terms of the
distribution of distances between the atoms and a set of pre-defined
points. This procedure characterizes the shape in the form of a 12-D
real valued fingerprint. As a result, the calculation procedure is
rapid and once calculated shape similarities can be evaluated rapidly
via a simple distance calculation. It should also be noted that shape
indices based on the molecular topology have also been described
\cite{Kier:1985aa}, which are very rapid to calculate (somewhat
analogous to the VSA descriptors) but do not directly represent the
molecular shape and instead infer it from topological connectivity.

@@

\textbf{Descriptors for inorganic compounds} Inorganic materials are
sufficiently different from small molecules that traditional
descriptors are not very useful. Recently, a number of descriptors for
inorganic materials are very similar to constitutional descriptors -
essentially counts of various crystal parameters such as accessible
surface areas and volumes
\cite{Willems:2012zr,Haranczyk:2010ys,Mackay:1984ve}. Other
descriptors characterize the periodic voids of porous materials via
tessellation methods \cite{Anurova:2009ly,Carr:2009kx}.

@@

The requirements listed here ensures that subsequent analyses
involving descriptors are robust. But in addition to numerical
robustness, it is important for many scenarios that the descriptor
have some form of physical interpretability.  For example, a model
that employs descriptors to predict cytotoxicity will be most useful if
it is numerically accurate as well as providing insight into which
structural features confer toxicity. In such a scenario, abstract
descriptors do not allow us to provide a physical explanation of the
models predictions. Another scenario is the ability to map descriptor
values back to the structure for visualization purposes such as the
``Glowing Molecule'' depictions of Segall et al
\cite{Segall:2009fk}. Such visualizations are practical only when
descriptor values can be related to structural features. Examples
include surface based descriptors such as CPSA's as well as explicit
structural descriptors such as paths, chains etc.

@@

Given the multiple implementations of many descriptors, it is natural
to compare their performance. As noted previously, it is desirable,
but unlikely, that implementations of the same descriptor using
different cheminformatics toolkits will be identical. Gupta et al
employed SMARTS based descriptors from the CDK and MOE to develop
decision tree models to predict human liver microsomal metabolic
stability \cite{Gupta:2010uq}. Their results indicated very similar
performance between the two implementations. In general, one would
expect a high degree of correlation between different implementations
of well-defined descriptors. Here ``well defined'' indicates that the
underlying parametrization or chemistry model is completely specified,
so that different implementations use them identically. For example,
correct implementations of the TPSA descriptor are expected to
correlate very well, and this is shown in Figure \ref{fig:tpsa} for
the ACD Labs and CDK implementations.

@@

\ctable[caption={A comparison of Topological Polar Surface Area values
generated using the CDK and ACD Labs software, for 57,857 molecules
taken from PubChem AID 1996.},cap={},label={fig:tpsa},botcap,figure]
{c}
{}
{
  \includegraphics[width=\linewidth]{tpsa-cdk-acd}
}
